---
title: "Best states data Analysis"
author: "Moreen Owino and Beatrice Kemboi"
date: "4/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load tidyverse}
library(tidyverse)
```
## R loading data on infrastructure: 
## metric Road quality

```{r, `read_data`}
road_quality_data <- read_csv("road_quality_data.csv")
road_quality_data

```


## filtering out unwanted rows
## renaming the states
## select the required columns

```{r, wrangling and renaming }
wrangled_data_roads <- road_quality_data %>% 
filter(grepl('Percent Acceptable', Class))%>%
rename(International_Roughness_Index = Value) %>%
select(Year,State,International_Roughness_Index ) 


wrangled_data_roads

```


#  Using pivot wider to tidy the data

```{r, pivot_wider}
final_road_data1 <- wrangled_data_roads %>%
pivot_wider(names_from = Year, values_from = International_Roughness_Index)

final_road_data1
```


# Deleting unwanted columns using dplyr select
# using mutate to get the sum and average

```{r, using mutate to get 10yr mean}
ten_year_roadmean <- final_road_data1 %>%
select(-c(12:19)) %>%
mutate("10year_mean" =rowMeans(select(.,`2011`,`2012`,`2013`,`2014`,`2015`,`2016`,`2017`,`2018`,`2019`,`2020`)))

ten_year_roadmean

```


```{r, using mutate to get mean per year}
rankedroad_data_2011to2020 <-ten_year_roadmean %>%
arrange(desc(`10year_mean`)) %>% 
mutate('2011_ranking' =min_rank(desc(`2011`))) %>% 
mutate('2012_ranking' = min_rank(desc(`2012`) )) %>% 
mutate('2013_ranking' = min_rank(desc(`2013`)) ) %>% 
mutate('2014_ranking' = min_rank(desc(`2014`)) ) %>% 
mutate('2015_ranking' = min_rank(desc(`2015`)) ) %>% 
mutate('2016_ranking' = min_rank(desc(`2016`)) ) %>% 
mutate('2017_ranking' = min_rank(desc(`2017`)) ) %>% 
mutate('2018_ranking' = min_rank(desc(`2018`)) ) %>% 
mutate('2019_ranking' = min_rank(desc(`2019`)) ) %>% 
mutate('2020_ranking' = min_rank(desc(`2020`)) ) %>% 
mutate('10year_ranking' = min_rank(desc(`10year_mean`)) )

rankedroad_data_2011to2020 

```

# summary of states and their ranks

```{r, summary state and rank}
roadrank_summary <- rankedroad_data_2011to2020  %>% 
select(State, c(`2011_ranking`:`10year_ranking`))

roadrank_summary

```

# Importing electricity prices per state

```{r, read electricity price data}
electricity_price_state <- read_csv("state_electricity_price.csv")

electricity_price_state

```


# data wrangling to clean up the data
# separating date column into year,month,day
# making the tibble wider
# splitting the date column into year, month, date


```{r, tidying electricity price data}
electricity_price <- electricity_price_state %>%
separate(date, into = c("year", "month","date"), sep = "-")%>%
select(-c(3:4)) %>%
pivot_wider(names_from = year, values_from = value )


   
electricity_price
```

# deleting unwanted rows
```{r}
final_electricity_price1 <- electricity_price %>%
select(-c(2:11, 22)) 

final_electricity_price1

```

```{r,  electricity price 10 year mean }
elec_tenyear_mean <- final_electricity_price1 %>%
mutate("10year_mean" =rowMeans(select(.,`2011`,`2012`,`2013`,`2014`,`2015`,`2016`,`2017`,`2018`,`2019`,`2020`)))

elec_tenyear_mean

```

```{r, using mutate to get mean per year}
elec_meandata_2011to2020 <- elec_tenyear_mean %>%
arrange(desc(`10year_mean`)) %>% 
mutate('2011_ranking' =min_rank(desc(`2011`))) %>% 
mutate('2012_ranking' = min_rank(desc(`2012`) )) %>% 
mutate('2013_ranking' = min_rank(desc(`2013`)) ) %>% 
mutate('2014_ranking' = min_rank(desc(`2014`)) ) %>% 
mutate('2015_ranking' = min_rank(desc(`2015`)) ) %>% 
mutate('2016_ranking' = min_rank(desc(`2016`)) ) %>% 
mutate('2017_ranking' = min_rank(desc(`2017`)) ) %>% 
mutate('2018_ranking' = min_rank(desc(`2018`)) ) %>% 
mutate('2019_ranking' = min_rank(desc(`2019`)) ) %>% 
mutate('2020_ranking' = min_rank(desc(`2020`)) ) %>% 
mutate('10year_ranking' = min_rank(desc(`10year_mean`)) )

elec_meandata_2011to2020

```

# summary of state and rankings 2011-2020

```{r}
elecprice_rank_summary <- elec_meandata_2011to2020 %>% 
select(state, c(`2011_ranking`:`10year_ranking`))

elecprice_rank_summary

```



# Crime rate data

```{r, reading_crime_data}
Corrections_data <- read_csv("Corrections data.csv")

Corrections_data
```


```{r, deleting_columns}
data_correction <- Corrections_data %>%
slice(-c(1:7)) 


data_correction
``` 
# continuing data deletion
# Rename the name of coloumn 1
# Separate coloumn 1 
# using the first row as the header

```{r}
data_correction2 <- data_correction %>%
slice(-c(56:735)) %>%
slice(-c(2:4)) %>%
rename(State = 1) %>%
separate(State, c("state"), extra = 'drop') %>%
select (-c(2:35)) %>%
purrr::set_names(as.character(slice(., 1))) %>%
slice(-1)


data_correction2

```


```{r}
final_data_correction <- data_correction2 %>%
rename(State = 1) %>%
mutate(sum = rowSums(across(where(is.numeric)))) %>%
mutate(average = round((sum / 9), digits = 2))

final_data_correction

```


# loading data on population 

```{r, loading data on population}
population <- read_csv("Statespopulation_2011to2020.csv") 

population

```
# rename column 1
```{r}
renamed_population <- population %>%
rename(State = 1)
renamed_population

```

# inner join 

```{r, add the population data to the data correction data}
combined_correction <- merge(x = final_data_correction, y = renamed_population, by = "State" )

combined_correction

```
# deleting unwanted columns in the combined data
```{r, deleting unwanted columns}
data_combined_correction <- combined_correction %>%
mutate(correction_2011 = `2011.x`/`2011.y`,
       correction_2012 = `2012.x` /`2012.y`,
       correction_2013 = `2013.x`/`2013.y`,
       correction_2014 = `2014.x` /`2014.y`,
       correction_2015 = `2015.x`/`2015.y`,
       correction_2016 = `2016.x` /`2016.y`,
       correction_2017 = `2017.x`/`2017.y`,
       correction_2018 = `2018.x`/`2018.y`,
       correction_2019 = `2019.x` /`2019.y`
)
       
data_combined_correction      

```
# selecting the correction rate per state

```{r, using select to select the correction data }
combined_correction_data <- data_combined_correction %>%
select(State,correction_2011:correction_2019) %>%
mutate("10year_mean" =rowMeans(select(.,`correction_2011`,`correction_2012`,`correction_2013`,`correction_2014`,`correction_2015`,`correction_2016`,`correction_2017`,`correction_2018`,`correction_2019`)))

combined_correction_data
```
# ranked correction data

```{r, ranked_correction_data}
ranked_correction_data <- combined_correction_data %>%
arrange(desc(`10year_mean`)) %>% 
mutate('2011_ranking' =min_rank(desc(`correction_2011`))) %>% 
mutate('2012_ranking' = min_rank(desc(`correction_2012`) )) %>% 
mutate('2013_ranking' = min_rank(desc(`correction_2013`)) ) %>% 
mutate('2014_ranking' = min_rank(desc(`correction_2014`)) ) %>% 
mutate('2015_ranking' = min_rank(desc(`correction_2015`)) ) %>% 
mutate('2016_ranking' = min_rank(desc(`correction_2016`)) ) %>% 
mutate('2017_ranking' = min_rank(desc(`correction_2017`)) ) %>% 
mutate('2018_ranking' = min_rank(desc(`correction_2018`)) ) %>% 
mutate('2019_ranking' = min_rank(desc(`correction_2019`)) ) %>% 
mutate('10year_ranking' = min_rank(desc(`10year_mean`)) )

ranked_correction_data
```

# ranked correction data
```{r}
final_ranked_correction <- ranked_correction_data %>%
select(State, 12:20)

final_ranked_correction
```

 

# EDUCATION DATA
#load data


```{r}
highschool_graduation <- read_csv("H_graduation_rate.csv")

highschool_graduation

```

# wrangle data
# 

```{r, deleting unwanted data}
graduation_rate <- highschool_graduation %>%
slice(-c(1:6, 58:70)) %>%
select(-c(3,5,7, 14:28))

graduation_rate
```

# Rename the header columns and assign Years
# Get sum and average using dplyr mutate
```{r, renaming columns }
graduation_rate2 <- graduation_rate %>%
rename(state = 1,
      "2011" = 2,
      "2012" = 3,
      "2013" = 4,
      "2014" = 5,
      "2015" = 6,
      "2016" = 7,
      "2017" = 8,
      "2018" = 9,
      "2019" = 10
      
       ) 

graduation_rate2

```

# Calculating mean values to replace NA's

```{r}
idaho_mean <- c(77,79,80,80,81,81)
mean_idaho <- (sum(idaho_mean))/6

kentucky_mean <- c(86,88,88,89,90,90,91)
mean_kentucky <- (sum(kentucky_mean))/7

Oklahoma_mean <- c(85,83,83,82,83,82,85)
mean_Oklahoma <- (sum(Oklahoma_mean))/7

```

# renaming cells with no value as NA


```{r, replacing_na} 
graduation_rate2 [13, 2:4] = '79'
graduation_rate2 [18, 2:3] = '88'
graduation_rate2 [37, 2:3] = '83'



```




# Ranking the average graduation rate over the years
 
```{r, 10year graduation rate mean}
graduation_rate2[,2:10] <- lapply(graduation_rate2[,2:10],as.numeric)

nineyear_graduation <- graduation_rate2 %>%
mutate("10year_mean" =rowMeans(select(., `2011`,`2012`,`2013`,`2014`,`2015`,`2016`,`2017`,`2018`,`2019`))
)
nineyear_graduation
```
# ranked_correction_data

```{r, ranked_correction_data}
ranked_graduation_data <- nineyear_graduation %>%
arrange(desc(`10year_mean`)) %>% 
mutate('2011_ranking' =min_rank(desc(`2011`))) %>% 
mutate('2012_ranking' = min_rank(desc(`2012`) )) %>% 
mutate('2013_ranking' = min_rank(desc(`2013`)) ) %>% 
mutate('2014_ranking' = min_rank(desc(`2014`)) ) %>% 
mutate('2015_ranking' = min_rank(desc(`2015`)) ) %>% 
mutate('2016_ranking' = min_rank(desc(`2016`)) ) %>% 
mutate('2017_ranking' = min_rank(desc(`2017`)) ) %>% 
mutate('2018_ranking' = min_rank(desc(`2018`)) ) %>% 
mutate('2019_ranking' = min_rank(desc(`2019`)) ) %>% 
mutate('10year_ranking' = min_rank(desc(`10year_mean`)) )

ranked_graduation_data

```

# final_ranked_data

```{r}
final_ranked_graduation <- ranked_graduation_data %>%
select(state, 12:21)

final_ranked_graduation
```




